\pkg{cubble}: An \proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2023-05-30 2-cubble.Rmd
subsection coercing from foreign objects

@@

### Create from separate spatial and temporal table

@@

make_cubble(spatial = stations, temporal = meteo_ts, 
            coords = c(long, lat))

make_cubble(spatial = stations_sf, temporal = meteo_ts)
```

### Coerce from other objects

  - import different data: an expand data, spatial + temporal, netcdf, stars, sftime

The function \code{as_cubble()} is used to create a \code{cubble} object with three arguments: \code{key} as the spatial identifier; \code{index} as the temporal identifier; and a vector of \code{coords} in the order (longitude, latitude). The arguments \code{key} and \code{index} follow the conventions in the \pkg{tsibble} package to describe the temporal order and multiple series while \code{coords} specifies the spatial location of each site. The code below creates a \code{cubble} object out of \code{climate_flat} (a single \code{tibble}) with \code{id} as the key, \code{date} as the index, and \code{c(long, lat)} as the coordinates:

@@

A \code{cubble} class can be created from various common spatio-temporal data formats, including \code{tibble}, \code{tsibble}, and \code{sf} when both spatial and temporal information are available. Section \@ref(covid) describes converting multiple tables into a \code{cubble} object and Section \@ref(netcdf) illustrates how to convert a NetCDF object.

*netcdf*

The NetCDF data is another format commonly used for storing spatio-temporal data. It has two main components: *dimension* for defining the spatio-temporal grid (longitude, latitude, and time) and *variable* that populates the defined grid. Attributes  can be associated with dimensions or variables. Because there can be many different styles of representing this information there is a metadata convention [@hassell_data_2017] to standardize the format of the attributes. A few packages in \proglang{R} exist for manipulating NetCDF data and these include a high-level R interface: \pkg{ncdf4} [@ncdf4], a low-level interface that calls a C-interface: \pkg{RNetCDF} [@rnetcdf], and a tidyverse implementation: \pkg{tidync} [@tidync].

The \pkg{cubble} package provides an \code{as_cubble()} method to coerce the \code{ncdf4} class from the \pkg{ncdf4} package into a \code{cubble} object. It maps each combination of longitude and latitude into an \code{id} as the \code{key}:

```{r echo = TRUE, eval = FALSE}
dt <- as_cubble(my_ncdf, vars = c("q", "z"))
```

Sometimes NetCDF data can be quite large, and it is best to subset the data when converting to a \code{cubble} object. We would recommend reducing to about 300 $\times$ 300 grid points for three daily variables in one year. A 300 by 300 spatial grid can be a bounding box of [100, -80, 180, 0] at 0.25 degree resolution or a global bounding box [-180, -90, 180, -90] at 1 degree resolution.
The size of spatial grid can be reduced if longer time periods or more variables are needed, through the arguments \code{long_range} and \code{lat_range}:

```{r eval = FALSE, echo = TRUE}
dt <- as_cubble(
  my_ncdf, vars = c("q", "z"),
  long_range = seq(-180, 180, 1), lat_range = seq(-90, 90, 1)
  )
```

*stars*

*sftimes*

>>>>

The arguments \code{key} and \code{index} follow the conventions in the \pkg{tsibble} package to describe the temporal order and multiple series while \code{coords} specifies the spatial location of each site.

@@

### Creating from separate spatial and temporal table

@@

make_cubble(spatial = stations, temporal = meteo_ts, 
            coords = c(long, lat))
```

```{r echo = TRUE}
make_cubble(spatial = stations_sf, temporal = meteo_ts)
```

### Coercing from foregin objects

Spatio-temporal data in other foreign objects can be coerced into cubble with the function \code{as_cubble()}. This includes casting from a \code{tibble} or \code{data.frame} with both spatial and temporal information, a netCDF object, a \code{stars} object [@stars], and a \code{sftime} object [@sftime]. The two examples below shows the casting from a tibble and a netcdf object to a cubble object.

The dataset \code{climate_flat} joins the spatial data, \code{stations}, with the temporal data, \code{meteo}, into a single tibble object and it can be coerced into a cubble using:

@@

The NetCDF data is another format commonly used for storing spatio-temporal data. In \proglang{R}, packages for wrangling NetCDF data include a high-level R interface: \pkg{ncdf4} [@ncdf4], a low-level interface that calls a C-interface: \pkg{RNetCDF} [@rnetcdf], and a tidyverse implementation: \pkg{tidync} [@tidync]. The code below casts a netCDF object in the \code{ncdf4} class into a cubble object:

```{r echo = TRUE}
path <- system.file("ncdf/era5-pressure.nc", package = "cubble")
raw <- ncdf4::nc_open(path)
as_cubble(raw)
```

Sometimes, one may want to read in a subset of the netcdf data and the argument \code{vars}, \code{long_range} and \code{lat_range} can be used to subset on the variable and the grid resolution:

```{r echo = TRUE, eval = FALSE}
as_cubble(
  raw, vars = "q",
  long_range = seq(-180, 180, 1), lat_range = seq(-90, 90, 1)
  )
```

We would recommend reducing to about 300 $\times$ 300 grid points for three daily variables in one year. A 300 by 300 spatial grid can be a bounding box of [100, -80, 180, 0] at 0.25 degree resolution or a global bounding box [-180, -90, 180, -90] at 1 degree resolution.

### 2023-05-30 jss/main.tex
subsection coercing from foreign objects

\hypertarget{create-from-separate-spatial-and-temporal-table}{%
\subsubsection{Create from separate spatial and temporal table}\label{create-from-separate-spatial-and-temporal-table}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> make_cubble(spatial = stations_sf, temporal = meteo, 
+             key = id, index = date)
R> 
R> make_cubble(spatial = stations, temporal = meteo_ts, 
+             coords = c(long, lat))
R> 
R> make_cubble(spatial = stations_sf, temporal = meteo_ts)
\end{CodeInput}
\end{CodeChunk}

\hypertarget{coerce-from-other-objects}{%
\subsubsection{Coerce from other objects}\label{coerce-from-other-objects}}

\begin{itemize}
\tightlist
\item
  import different data: an expand data, spatial + temporal, netcdf, stars, sftime
\end{itemize}

The function \code{as_cubble()} is used to create a \code{cubble} object with three arguments: \code{key} as the spatial identifier; \code{index} as the temporal identifier; and a vector of \code{coords} in the order (longitude, latitude). The arguments \code{key} and \code{index} follow the conventions in the \pkg{tsibble} package to describe the temporal order and multiple series while \code{coords} specifies the spatial location of each site. The code below creates a \code{cubble} object out of \code{climate_flat} (a single \code{tibble}) with \code{id} as the key, \code{date} as the index, and \code{c(long, lat)} as the coordinates:

@@

A \code{cubble} class can be created from various common spatio-temporal data formats, including \code{tibble}, \code{tsibble}, and \code{sf} when both spatial and temporal information are available. Section \ref{covid} describes converting multiple tables into a \code{cubble} object and Section \ref{netcdf} illustrates how to convert a NetCDF object.

\emph{netcdf}

The NetCDF data is another format commonly used for storing spatio-temporal data. It has two main components: \emph{dimension} for defining the spatio-temporal grid (longitude, latitude, and time) and \emph{variable} that populates the defined grid. Attributes can be associated with dimensions or variables. Because there can be many different styles of representing this information there is a metadata convention \citep{hassell_data_2017} to standardize the format of the attributes. A few packages in \proglang{R} exist for manipulating NetCDF data and these include a high-level R interface: \pkg{ncdf4} \citep{ncdf4}, a low-level interface that calls a C-interface: \pkg{RNetCDF} \citep{rnetcdf}, and a tidyverse implementation: \pkg{tidync} \citep{tidync}.

The \pkg{cubble} package provides an \code{as_cubble()} method to coerce the \code{ncdf4} class from the \pkg{ncdf4} package into a \code{cubble} object. It maps each combination of longitude and latitude into an \code{id} as the \code{key}:

\begin{CodeChunk}
\begin{CodeInput}
R> dt <- as_cubble(my_ncdf, vars = c("q", "z"))
\end{CodeInput}
\end{CodeChunk}

Sometimes NetCDF data can be quite large, and it is best to subset the data when converting to a \code{cubble} object. We would recommend reducing to about 300 \(\times\) 300 grid points for three daily variables in one year. A 300 by 300 spatial grid can be a bounding box of {[}100, -80, 180, 0{]} at 0.25 degree resolution or a global bounding box {[}-180, -90, 180, -90{]} at 1 degree resolution.
The size of spatial grid can be reduced if longer time periods or more variables are needed, through the arguments \code{long_range} and \code{lat_range}:

\begin{CodeChunk}
\begin{CodeInput}
R> dt <- as_cubble(
+   my_ncdf, vars = c("q", "z"),
+   long_range = seq(-180, 180, 1), lat_range = seq(-90, 90, 1)
+   )
\end{CodeInput}
\end{CodeChunk}

\emph{stars}

\emph{sftimes}

>>>>

\hypertarget{creating-from-separate-spatial-and-temporal-table}{%
\subsubsection{Creating from separate spatial and temporal table}\label{creating-from-separate-spatial-and-temporal-table}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> make_cubble(spatial = stations_sf, temporal = meteo, 
+             key = id, index = date)
R> 
R> make_cubble(spatial = stations, temporal = meteo_ts, 
+             coords = c(long, lat))
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> make_cubble(spatial = stations_sf, temporal = meteo_ts)
\end{CodeInput}
\begin{CodeOutput}
# cubble:   key: id [3], index: date, nested form, [sf]
# extent:   [144.8321, -37.98, 145.0964, -37.6655]
# temporal: prcp [dbl], tmax [dbl], tmin [dbl]
  id           long   lat  elev name  wmo_id            geometry
  <chr>       <dbl> <dbl> <dbl> <chr>  <dbl>         <POINT [°]>
1 ASN00086038  145. -37.7  78.4 esse~  95866 (144.9066 -37.7276)
2 ASN00086077  145. -38.0  12.1 moor~  94870   (145.0964 -37.98)
3 ASN00086282  145. -37.7 113.  melb~  94866 (144.8321 -37.6655)
# i 1 more variable: ts <list>
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{coercing-from-foregin-objects}{%
\subsubsection{Coercing from foregin objects}\label{coercing-from-foregin-objects}}

Spatio-temporal data in other foreign objects can be coerced into cubble with the function \code{as_cubble()}. This includes casting from a \code{tibble} or \code{data.frame} with both spatial and temporal information, a netCDF object, a \code{stars} object \citep{stars}, and a \code{sftime} object \citep{sftime}. The two examples below shows the casting from a tibble and a netcdf object to a cubble object.

The dataset \code{climate_flat} joins the spatial data, \code{stations}, with the temporal data, \code{meteo}, into a single tibble object and it can be coerced into a cubble using:

@@

The NetCDF data is another format commonly used for storing spatio-temporal data. In \proglang{R}, packages for wrangling NetCDF data include a high-level R interface: \pkg{ncdf4} \citep{ncdf4}, a low-level interface that calls a C-interface: \pkg{RNetCDF} \citep{rnetcdf}, and a tidyverse implementation: \pkg{tidync} \citep{tidync}. The code below casts a netCDF object in the \code{ncdf4} class into a cubble object:

\begin{CodeChunk}
\begin{CodeInput}
R> path <- system.file("ncdf/era5-pressure.nc", package = "cubble")
R> raw <- ncdf4::nc_open(path)
R> as_cubble(raw)
\end{CodeInput}
\begin{CodeOutput}
# cubble:   key: id [26565], index: time, nested form
# extent:   [113, -53, 153, -12]
# temporal: q [dbl], z [dbl]
     id  long   lat ts              
  <int> <dbl> <dbl> <list>          
1     1  113    -12 <tibble [6 x 3]>
2     2  113.   -12 <tibble [6 x 3]>
3     3  114.   -12 <tibble [6 x 3]>
4     4  114.   -12 <tibble [6 x 3]>
5     5  114    -12 <tibble [6 x 3]>
# i 26,560 more rows
\end{CodeOutput}
\end{CodeChunk}

Sometimes, one may want to read in a subset of the netcdf data and the argument \code{vars}, \code{long_range} and \code{lat_range} can be used to subset on the variable and the grid resolution:

\begin{CodeChunk}
\begin{CodeInput}
R> as_cubble(
+   raw, vars = "q",
+   long_range = seq(-180, 180, 1), lat_range = seq(-90, 90, 1)
+   )
\end{CodeInput}
\end{CodeChunk}

We would recommend reducing to about 300 \(\times\) 300 grid points for three daily variables in one year. A 300 by 300 spatial grid can be a bounding box of {[}100, -80, 180, 0{]} at 0.25 degree resolution or a global bounding box {[}-180, -90, 180, -90{]} at 1 degree resolution.
